---
title: "Assignment 5: Data Visualization"
author: "Judy Hua Zhu"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
library(ggplot2)
here()

processed_data = "./Data/Processed_KEY"

PeterPaul.chem.nutrients <- read.csv(
  here(processed_data,"NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE)

Litter <- read.csv(
  here(processed_data,"NEON_NIWO_Litter_mass_trap_Processed.csv"), 
  stringsAsFactors = TRUE)

#2 
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate)
Litter$collectDate <- as.Date(Litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
my_theme <- theme_bw() + 
  theme(
    line = element_line(color='purple',linewidth = 2),
    axis.text = element_text(color = "black"), 
    legend.position = "right",
    title.position = "top",
    plot.title.size = 2,
    points.size = 0.5,
    panel.grid.major =  element_line(linewidth=0.5),
    panel.grid.minor =  element_line(linewidth=0.5)
    )


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

Q4 <- ggplot(PeterPaul.chem.nutrients, aes(x = po4, y = tp_ug, 
                                           color = lakename)) +
  geom_point() +                                # Scatter plot
  geom_smooth(method = "lm", se = FALSE) +      # Line of best fit without confidence interval shading
  xlim(0, 50) +                                 # Adjust x-axis limits to hide extreme values
  ylim(0, 150) +                                # Adjust y-axis limits to hide extreme values
  labs(title = "Total Phosphorus vs. Phosphate by Lake",
       x = "Phosphate (po4)",
       y = "Total Phosphorus (tp_ug)",
       color = "Lake name") +
  my_theme 

Q4
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots} 

#5 
PeterPaul.chem.nutrients$month <- factor(PeterPaul.chem.nutrients$month,
       levels=1:12,
       labels = month.abb)

Q5_temp <- ggplot(PeterPaul.chem.nutrients, aes(color=lakename, x= month))+
  geom_boxplot (aes(y = temperature_C))+
  theme(legend.position = "none",
        axis.title.x = element_blank())+
  labs(title = "Temperature, TN and TP by month")

Q5_tp <- ggplot(PeterPaul.chem.nutrients, aes(color=lakename, x= month))+
  geom_boxplot(aes(y = tp_ug)) +
  theme(axis.title.x = element_blank())+
  labs(color = "Lake name")

Q5_tn <- ggplot(PeterPaul.chem.nutrients, aes(color=lakename, x= month))+
  geom_boxplot(aes(y = tn_ug))+
  theme(legend.position = "none")

plot_grid (Q5_temp, Q5_tp, Q5_tn, nrow = 3, align = "v")
        
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Peter lake has lower summer temperature but higher TP and TN values than Paul lake. Also, TN and TP is a bit higher in summer than spring and fall, which are the only data available in this dataset.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6 Date not shown in x-axis; 
Q6 <- Litter %>%
  filter (functionalGroup=="Needles") %>% 
  ggplot (aes(x = collectDate, y = dryMass, color= nlcdClass)) +
  geom_point()+                               
  ylim(0, 50) +
  scale_x_date(date_labels = "%Y")+
  ylab("Dry Mass")+
  xlab("Collect Date") +
  labs(title = "Dry Mass of Needles in different NLCD",color = "NLCD Class")

Q6

#7
Q7 <- Litter %>%
  filter (functionalGroup=="Needles") %>% 
  ggplot (aes(x = collectDate, y = dryMass)) +
  geom_point()+                               
  ylim(0, 50) +
  scale_x_date(date_labels = "%Y")+
  ylab("Dry Mass")+
  xlab("Collect Date") +
  labs(title = "Dry Mass of Needles in different NLCD",color = "NLCD Class")+
  facet_wrap(vars(nlcdClass),ncol = 3) +
  my_theme

Q7
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think plot 7 is more effective, as it shows the comparison of data in different NLCD more clearly by separating the data. Plot 6 is hard to compare the NLCDs since the data is cluttered together and is hard for color-weak or even normal people. However, if we care more on the trend of dry mass change with time instead of the NLCD class, plot 6 is a good represent.